Constructing Biological Regulatory Networks from Process Hitting models \\\rewriteil{Or possibly: “Complementarity between PH, IG and Thomas modeling”}

### 2013-06-18 parts/infer-ig.tex
finish new inference

Finally, we infer that $a$ has a self-influence if its current level can have an impact on its own
evolution in a given configuration $\sigma$ of any group of influence on $a$.
Given $a_i,a_{i+1}\in L_a$, we pick $a_j\in\focals(a,\{a_i\},\ctx_g(\sigma\{a_i\}))$ and
$a_k\in\focals(a,\{a_{i+1}\},\ctx_g(\sigma\{a_{i+1}\}))$.
If $k=j+1$, we can not conclude as there is no difference in the evolution of both levels.
If $k\neq j+1$ and $k-j\neq 0$, then $a_i$ and $a_{i+1}$ have divergent evolutions: we infer an
influence of sign of $k-j$ at threshold $i+1$.
We note that some aspects of this inference are arbitrary and may impact the number of parameters to
infer in the next section.
In particular, in some cases, the use of intervals for Thomas' parameters drops the requirement of
inferring a self-activation.

\pref{pps:inference-edges} details the inference of all existing influences between components occurring
with a threshold $t$.
These influences are split into positive and negative ones, and represent possible edges in the final IG.
We do not consider the cases where a component has no visible influence on another.
\begin{proposition}[Edges inference]\label{pps:inference-edges}
We define the set of positive (resp. negative) influences $\hat{E}_+$ (resp. $\hat{E}_-$) for any
$a\in\Gamma$ by:
\begin{align}
\begin{split}
\forall b\in\PHpredecgene{a}, b &\neq a, \\
b\xrightarrow t a \in \hat{E}_s & \Longleftrightarrow
  \exists g \in X(a), b \in g, 
  \exists \sigma\in\textstyle\prod_{c\in g \cup \{ a \}} L_c, \exists b_i,b_{i+1}\in \PHl_b,\\
& \qquad\qquad
        \exists a_{j}\in\focals(a, \{\sigma[a]\}, \ctx_b(\sigma\{b_i\})), \\
& \qquad\qquad
        \exists a_{k}\in\focals(a, \{\sigma[a]\}, \ctx_b(\sigma\{b_{i+1}\})), \\
& \qquad\qquad\qquad
                        s = \f{sign}(k - j) \wedge t = i+1
\end{split}
\label{eq:edges-inference-b}
\\
\begin{split}
a \xrightarrow t a \in \hat{E}_s & \Longleftrightarrow
\exists g \in X(a), \sigma \in \textstyle\prod_{b\in g \cup \{ a \}} L_b,
      \exists a_i,a_{i+1}\in \PHl_a, \\
& \qquad\qquad
        \exists a_{j}\in\focals(a, \{a_i\}, \ctx_g(\sigma\{a_i\})), \\
& \qquad\qquad
        \exists a_{k}\in\focals(a, \{a_{i+1}\},  \ctx_g(\sigma\{a_{i+1}\})), \\
& \qquad\qquad\quad
      k \neq j+1
        \wedge s = \f{sign}(k - j) \wedge t = i+1
\end{split}
\label{eq:edges-inference-a}
\end{align}
where $s \in \{ +, - \}$,
$\f{sign}(n) = + \EQDEF n > 0$,
$\f{sign}(n) = - \EQDEF n < 0$,
and $\f{sign}(0) \DEF 0$.
\end{proposition}

@@

\todo{Add a new subsection about the alternative inference (cf.~A.~Richard)}

@@

This second inference is complementary to the one presented in the previous subsection.
It is based on an approach considering the global dynamics of a component rather than independent dynamics of each regulator;
in that, it is analogous to standard IG inferences from discrete maps \cite{RiCo07}.
Contrary to the previously presented method, we focus on a configuration (or: sub-state) $\sigma$
on the whole set $\reg{a}$ of predecessors of a component $a$ (rather than on a subset of cooperating regulators).

\rewriteil{$B_a(\sigma)$ = Ensemble des processus vers lesquels on bondit}

\begin{align}
\begin{split}\label{eq:edges-inference-richard-pos}
  \forall b\in\PHpredecgene{a}, \qquad\qquad\\
  b\xrightarrow{t+1} a \in \hat{E}_+ \Longleftrightarrow\ &
    \exists a_j \in B_a(\sigma\{b_{t}\}), \exists a_k \in B_a(\sigma\{b_{t+1}\}),\\
    &\qquad a_j \leq \PHget{\sigma}{a} < a_k \warnul{\vee a_j < \PHget{\sigma}{a} \leq a_k \quad^*}
\end{split}\\
\begin{split}\label{eq:edges-inference-richard-neg}
  \forall b\in\PHpredecgene{a}, \qquad\qquad\\
  b\xrightarrow{t+1} a \in \hat{E}_- \Longleftrightarrow\ &
    \exists a_j \in B_a(\sigma\{b_{t}\}), \exists a_k \in B_a(\sigma\{b_{t+1}\}),\\
    &\qquad a_j > \PHget{\sigma}{a} \geq a_k \warnul{\vee a_j \geq \PHget{\sigma}{a} > a_k \quad^*}
\end{split}
\end{align}

\todo{\underline{$^*$Garder les conditions larges ?}}

>>>>

Finally, we infer that $a$ has a self-influence if its current level can have an impact on its own
evolution in a given configuration $\sigma$ of any group of influence on $a$ (\pref{eq:edges-inference-a}).
Given $a_i,a_{i+1}\in L_a$, we pick $a_j\in\focals(a,\{a_i\},\ctx_g(\sigma\{a_i\}))$ and
$a_k\in\focals(a,\{a_{i+1}\},\ctx_g(\sigma\{a_{i+1}\}))$.
If $k=j+1$, we can not conclude as there is no difference in the evolution of both levels.
If $k\neq j+1$ and $k-j\neq 0$, then $a_i$ and $a_{i+1}$ have divergent evolutions: we infer an
influence of sign of $k-j$ at threshold $i+1$.
We note that some aspects of this inference are arbitrary and may impact the number of parameters to
infer in the next section.
In particular, in some cases, the use of intervals for Thomas' parameters drops the requirement of
inferring a self-activation.

\pref{pps:inference-edges} details the inference of all existing influences between components occurring
with a threshold $t$.
These influences are split into positive and negative ones, and represent possible edges in the final IG.
We do not consider the cases where a component has no visible influence on another.
\begin{proposition}[Influences inference]\label{pps:inference-edges}
We define the set of positive (resp. negative) influences $\hat{E}_+$ (resp. $\hat{E}_-$) for any
$a\in\Gamma$ by:
\begin{align}
\begin{split}
\forall b\in\PHpredecgene{a}, b &\neq a, \\
b\xrightarrow t a \in \hat{E}_s & \Longleftrightarrow
  \exists g \in X(a), b \in g, 
  \exists \sigma\in\textstyle\prod_{c\in g \cup \{ a \}} \PHl_c, \exists b_i,b_{i+1}\in \PHl_b,\\
& \qquad\qquad
        \exists a_{j}\in\focals(a, \{\sigma[a]\}, \ctx_b(\sigma\{b_i\})), \\
& \qquad\qquad
        \exists a_{k}\in\focals(a, \{\sigma[a]\}, \ctx_b(\sigma\{b_{i+1}\})), \\
& \qquad\qquad\qquad
                        s = \f{sign}(k - j) \wedge t = i+1
\end{split}
\label{eq:edges-inference-b}
\\
\begin{split}
a \xrightarrow t a \in \hat{E}_s & \Longleftrightarrow
\exists g \in X(a), \sigma \in \textstyle\prod_{b\in g \cup \{ a \}} \PHl_b,
      \exists a_i,a_{i+1}\in \PHl_a, \\
& \qquad\qquad
        \exists a_{j}\in\focals(a, \{a_i\}, \ctx_g(\sigma\{a_i\})), \\
& \qquad\qquad
        \exists a_{k}\in\focals(a, \{a_{i+1}\},  \ctx_g(\sigma\{a_{i+1}\})), \\
& \qquad\qquad\quad
      k \neq j+1
        \wedge s = \f{sign}(k - j) \wedge t = i+1
\end{split}
\label{eq:edges-inference-a}
\end{align}
where $s \in \{ +, - \}$,
$\f{sign}(n) = + \EQDEF n > 0$,
$\f{sign}(n) = - \EQDEF n < 0$,
and $\f{sign}(0) \DEF 0$.
\end{proposition}

@@

@@

This second inference is complementary to the one presented in the previous subsection.
It is based on an approach considering the global dynamics of a component rather than independent dynamics of each regulator,
%in that, it is analogous to standard IG inferences from discrete maps \cite{RiCo07}.
and is inspired from the definition of Interaction Graph built from a Transition Graph as described in~\cite{Richard2010378}.
Contrary to the previously presented method, we focus on a configuration (or: sub-state) $\sigma$
on the whole set $\reg{a}$ of predecessors of a component $a$ (rather than on a subset of cooperating regulators).

For a given configuration $\sigma$ of a component $a$ and its regulators,
$\irB_a(\sigma)$ is the set of processes towards which $a$ can bounce (\pref{eq:possible-bounces}).
This set is defined using the set $\irF_a(\sigma)$ of action hitting $\PHget{\sigma}{a}$ in $\sigma$ (\pref{eq:possible-actions})
and the set of processes $\ctx$ hitting any process of $a$ (\pref{eq:ctx-sigma}).
If $a$ cannot be hit by any action in $\sigma$, then $\irB_a(\sigma) = \{ \PHget{\sigma}{a} \}$.
%
\begin{align}
\begin{split}\label{eq:possible-actions}
  \forall \sigma &\in \textstyle\prod_{c\in \reg{a} \cup \{ a \}} \PHl_c, \sigma' = \ctx_{\reg{a}}(\sigma),\\
  & \irF_a(\sigma) \DEF \{ a_k \in \PHl_a \mid \exists b \in \PHs, \exists \PHfrappe{b_i}{a_j}{a_k} \in \PHa,\\
  & \qquad\qquad\qquad\qquad\qquad \PHget{\sigma'}{b} = b_i \wedge \PHget{\sigma'}{a} = a_j \}\\
\end{split}
\end{align}
%
\begin{align}
\begin{split}\label{eq:possible-bounces}
  \forall \sigma &\in \textstyle\prod_{c\in \reg{a} \cup \{ a \}} \PHl_c,
%    \sigma' = \ctx_{\reg{a}}(\sigma),\\
  \irB_a(\sigma) \DEF 
  \begin{cases}
    \irF_a(\sigma)
      & \text{ if } \irF_a(\sigma) \neq \emptyset\\
    \{ \PHget{\sigma}{a} \}
      & \text{ if } \irF_a(\sigma) = \emptyset
  \end{cases}
\end{split}
\end{align}

The main idea of the inference detailed in \pref{pps:inference-edges-dynamics} is close to the one of \pref{ssec:infer-IG-structure}:
the presence of a positive (negative) influence of a component $b$ on $a$ denotes the fact that
there exists a state in which increasing the level of $b$ tends to make the future level of $a$ rise (drop) (\pref{eq:edges-inference-richard-pos} \& \eqref{eq:edges-inference-richard-neg}).
the main difference lies in the fact that this inference focuses on local evolutions (the possibility for $a$ to bounce away from its original state)
rather than global evolutions (by considering focal processes denoting the overall influence).
Therefore, an influence is inferred if the possible bounces of $a$ in $\sigma$ depend on the level of $b$.
%
\begin{proposition}[Influences inference]\label{pps:inference-edges-dynamics}
We define the set of positive (resp. negative) influences $\hat{E}_+$ (resp. $\hat{E}_-$) for any $a\in\Gamma$ by:
\begin{align}
\begin{split}\label{eq:edges-inference-richard-pos}
  \forall b\in\PHpredecgene{a}, \qquad\qquad\quad\\
  b\xrightarrow{t+1} a \in \hat{E}_+ \Longleftrightarrow\ &
    \exists \sigma \in \textstyle\prod_{c\in \reg{a} \cup \{ a \}} \PHl_c, \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}),\\
    &\qquad a_j \leq \PHget{\sigma}{a} < a_k \warnul{\vee a_j < \PHget{\sigma}{a} \leq a_k \quad^*}
\end{split}\\
\begin{split}\label{eq:edges-inference-richard-neg}
  \forall b\in\PHpredecgene{a}, \qquad\qquad\quad\\
  b\xrightarrow{t+1} a \in \hat{E}_- \Longleftrightarrow\ &
    \exists \sigma \in \textstyle\prod_{c\in \reg{a} \cup \{ a \}} \PHl_c, \{ b_t, b_{t+1} \} \subset \PHl_b \wedge b_t \in \sigma,\\
    &\qquad \exists a_j \in \irB_a(\sigma), \exists a_k \in \irB_a(\sigma\{b_{t+1}\}),\\
    &\qquad a_j > \PHget{\sigma}{a} \geq a_k \warnul{\vee a_j \geq \PHget{\sigma}{a} > a_k \quad^*}
\end{split}
\end{align}
\end{proposition}
%
\todo{\underline{$^*$Keep these large conditions?}}

Finally it is possible to infer the final IG by using the same reasoning as \pref{pps:inference-IG}:
a positive (negative) edge in inferred if there exist positives (negative) influences only, otherwise an unsigned edge in inferred.
We point out that the main use of this alternative inference is the fact that the conditions used to infer each influence
in \pref{eq:edges-inference-richard-pos} \& \eqref{eq:edges-inference-richard-neg}
allows to apply some results about the links between the structure of an IG and the dynamics of the model.
These results consist mainly in Thomas' conjectures, that focus on the presence of positive and negative circuits in the IG, and state that:
\begin{itemize}
  \item the presence of a positive circuit in the IG is a necessary condition for several fixed points,
  \item the presence of a negative circuit in the IG is a necessary condition for sustained oscillations.
\end{itemize}
The sign of a circuit is defined by the number of negative edges it contains;
given the nature of unsigned edges, the presence of such an edge in a circuit would count for the presence of both a positive and negative circuit.
These conjectures have been demonstrated in~\cite{Richard2010378} \towrite{Other references! (multi-stationarity \& boolean models)}.

\todo{Example}
